---
title: "Quick Start Guide to SplitR"
author: "Richard Iannone"
date: "June 19, 2014"
output: html_document
---

SplitR
======

SplitR is an R package that's great for conducting trajectory and dispersion modelling with [NOAA HYSPLIT](http://ready.arl.noaa.gov/HYSPLIT.php).

## Installation

Easy installation of SplitR from GitHub is entirely possible with the `devtools` package:

```{r}
# require(devtools)
# install_github('SplitR', 'rich-iannone')
```

## Description

SplitR allows you set up and run HYSPLIT in a very fast, easy, and organized manner. You can set up a few or, perhaps, thousands of trajectory or dispersion runs by using a single function. Because SplitR is an R interface to HYSPLIT, we can store output in data frames and take advantage of the vast selection of R packages to perform cluster analyses, generate and save plots, and more. This package simplifies the process of running HYSPLIT models by automating the downloading and storage of meteorological data files, providing a simple means to initiate a range of runs, and by providing outputs that can be easily applied to statistical analyses.

This package is in active development but is quite usable at present. Some of its features are:

- simple setup of trajectory and dispersion model runs (forward or backward)
- automated retrieval of meteorological data files
- organization of modelling presets and model output
- numerous batching options: set up models to run continuously with multiple start times throughout a single year, several years, or a predefined time range
- trajectory plotting and visualization of particle positions in dispersion runs

## Setup

Use of SplitR requires a basic installation of HYSPLIT, as obtained from the Air Resources Laboratory (ARL) of the National Oceanic and Atmospheric Administration (NOAA) [READY website](http://ready.arl.noaa.gov/HYSPLIT.php). 

For Windows installations, there are both [registered](http://ready.arl.noaa.gov/hyreg/HYSPLIT_pchysplit.php) and [unregistered](http://ready.arl.noaa.gov/HYSPLIT_hytrial.php) versions of HYSPLIT. The only difference in functionality between the two is that the unregistered version cannot perform model runs with forecast meteorological data. SplitR has the capability to perform forecast model runs so obtain the registered version if you intend to use that functionality. Regardless of version, both come with a standard installer that will place the main HYSPLIT folder (named 'hysplit') in the location of your choice. Within this main folder, there are 15 subdirectories.

For a Mac OS X installation, there is a [single distribution](http://ready.arl.noaa.gov/hyreg/HYSPLIT_applehysp.php) that has all functionality included.

Take note of the paths for the HYSPLIT executables and the working directory (in the subfolders 'exec' and 'working'). You can optionally create additional subfolders for locating HYSPLIT-ready meteorological data files, or, opt to place those in the HYSPLIT working directory.

## Example: HYSPLIT Trajectory Runs

To perform a series HYSPLIT trajectory model runs, use the SplitR `hysplit.trajectory` function:

```{r}
require(SplitR)

hysplit.trajectory(traj_name = "third",
                   start_lat_deg = 49.5, start_long_deg = -125.30364,
                   start_height_m_AGL = 5, simulation_duration_h = 24,
                   backtrajectory = TRUE, met_type = "gdas1",
                   vertical_motion_option = 0,
                   top_of_model_domain_m = 20000,
                   run_type = "range",
                   run_range = c("2012-03-15", "2012-03-16"),
                   daily_hours_to_start = c("00", "03", "06", "09", "12", "15", "18", "21"),
                   path_met_files = "/Users/riannone/Documents/SplitR/Met/",
                   path_output_files = "/Users/riannone/Documents/SplitR/Output/",
                   path_wd = "/Users/riannone/Documents/SplitR/Working/",
                   path_executable ="/Users/riannone/Documents/SplitR/Exec/")
```

This use of `hysplit.trajectory` sets up four trajectory runs that start at 00:00, 06:00, 12:00, and 18:00 UTC on March 12, 2012. These initial times are set using `run_type = "day"`, `run_day = "2012-03-12"`, and `daily_hours_to_start = c("00", "06", "12", "18")`. The model runs are forward runs (moving forward in time, set here using `backtrajectory = FALSE`) and not backtrajectory runs (set with `backtrajectory = TRUE`). These runs are 24 h in duration (`simulation_duration_h = 24`). The starting location of 42.83752ºN and 80.30364ºW is set using `start_lat_deg = 42.83752` and `start_long_deg = -80.30364`; the starting height of 5 m above ground level is set by `start_height_m_AGL = 5`. The meteorological options include the type of met data to use (1º GDAS data is used here with `met_type = "gdas1"`--there is also the option to use NCEP reanalysis data with the `met_type = "reanalysis"` setting), the vertical motion option (here, set as `vertical_motion_option = 0` which instructs HYSPLIT to use the vertical motion available in the met data files), and, the top of the model domain (set as 20,000 meters with `top_of_model_domain_m = 20000`). Four paths require specification:

- path to the meteorological data files (`path_met_files`)
- path to the output files (`path_output_files`)
- path to the working directory (`path_wd`)
- path to the executable directory, specifically here to that path containing the `hyts_std` executable (`path_executable`)

All paths should exist (i.e., SplitR won't create directories) and the paths provided in the above `hysplit.trajectory()` example should serve as examples for these path settings. If running in a Mac or Linux environment, use appropriate paths with forward slashes (paths using a tilde are acceptable).

The necessary meteorological data files relevant to the period being modelled will be downloaded from the NOAA FTP server (arlftp.arlhq.noaa.gov) if they are not present in the directory specified as the `path_met_files` argument. Note that SplitR does not currently provide an option to automatically delete these downloaded data files after the relevant model runs have been completed, so, keep in mind that available disk space may be issue with longer sequences of model runs (e.g., a GDAS1 met file for a week-long period can take up to 600 MB of disk space).

After this, four files should be generated:

- `traj(forward)-12-03-12-00-lat_42.83752_long_-80.30364-height_5-24h`
- `traj(forward)-12-03-12-06-lat_42.83752_long_-80.30364-height_5-24h`
- `traj(forward)-12-03-12-12-lat_42.83752_long_-80.30364-height_5-24h`
- `traj(forward)-12-03-12-18-lat_42.83752_long_-80.30364-height_5-24h`

On Mac/Linux, these files will be associated with a .zip archive that is named according to the value of `traj_name` (if provided) and the date/time of execution. The location of the archive will be that of the path provided in the `path_output_files` argument.

A data frame can be generated from these output files using the SplitR `trajectory.read` function:

```{r}
trajectory.df <- trajectory.read(archive_folder = "/Users/riannone/Documents/SplitR/Output/",
                                 year = NULL,
                                 start_height_m_AGL = NULL)
```

With this data frame, statistical analyses for the trajectories can be generated (e.g., average heights of trajectories after specified time periods, etc.). Furthermore, the `trajectory.df` data frame is fully compatible with the excellent 'openair' package that is available on CRAN. Plotting of the trajectory output data frame requires use of openair's `trajPlot` function:

```{r}
trajPlot(trajectory.df, map.fill = FALSE)
```

